---
title: "srm-overview: Social Relations Models for Network Analysis"
author: "Cassy Dorff, Shahryar Minhas, and Tosin Salau"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{srm-Overview: Social Relations Models for Network Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Package Overview

This vignette provides an overview of the `srm` package, which provides tools for estimating social relations models for network data. It breaks down and analyzes networks into nodal and dyadic dependencies and allows users to understand actor, partner, and dyadic effects in network data (Dorff and Ward, 2013; Dorff and Minhas, 2017).

**What users can do:**

(1) **Identify the most active actors in the network** - both senders and receivers patterns. Users can identify which actors are more active than expected after accounting for overall network structure and partner characteristics.

(2) **Identify overall network patterns** where users can separate systematic actor behaviors from relationship-specific effects.

(3) **Generate specific summary statistics for the network data** that calculates neabs and covariances that can reveal whether active actors are also active recievers

(4) **Create  visualizations** that separate actor behaviors from relationship-specific effects.

**In comparison to `netify`**, which provides general network statistics and visualizations, the srm package helps users break down network relationships allowing for deeper analysis. With `srm`, users are able to identify which actors are more or less active than typical and tells us why those connections are stronger or weaker than expected. While `netify` focuses on overall network patterns, srm centers on explaining the individual behaviors and relationships that drive those patterns.


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  fig.width = 6,
  fig.height = 4,
  dpi = 200,
  out.width = "100%",
  fig.align = "center",
  dev = "png" 
)

```


## Install and load Packages
```{r setup, eval=TRUE, warning=FALSE, message=FALSE, results='hide'}

# Install required packages if not already installed
if(!'devtools' %in% rownames(installed.packages())){
  install.packages('devtools', repos='https://cloud.r-project.org')
}
if(!'netify' %in% rownames(installed.packages())){
  install.packages('netify', repos='https://cloud.r-project.org')
}
if(!'tidyverse' %in% rownames(installed.packages())){
  install.packages('tidyverse', repos='https://cloud.r-project.org')
}
if(!'ggplot2' %in% rownames(installed.packages())){
  install.packages('ggplot2', repos='https://cloud.r-project.org')
}

# Install the srm package from GitHub if not already installed
if(!'srm' %in% rownames(installed.packages())){
  devtools::install_github("netify-dev/srm")
}

# Load all packages
library(devtools)
library(netify)
library(tidyverse)
library(ggplot2)
library(srm)
```

## Prepare Data

We will make use of data from ATOP (Alliance Treaty Obligations and Provisions) to demonstrate the how to use the `srm` package. The data can be downloaded at http://www.atopdata.org. For this tutorial, we have selected the atop5_1dy dataset and focus on military security consultations between countries from 2010 to 2018 where at least one partner is from the following 11 East Asian countries:  Japan, South Korea, Philippines, Thailand, Singapore, Malaysia, Indonesia, India, China, Vietnam, and North Korea.


```{r, echo=FALSE}

data("atop_EA")  # Load dataset

```


We will make use of the `netify` package to make a network object from the atop data. We will create 2 network objects, one for cross-sectional data and the other for longitudinal data as examples for this vignette. 

```{r, warning=FALSE, message=FALSE}
# cross-sectional
atop_EA_short <- netify::netify(
  dyad_data = atop_EA,
  actor1 = 'country1',
  actor2 = 'country2',
  symmetric = TRUE,
  sum_dyads = TRUE,
  diag_to_NA = TRUE,
  missing_to_zero = TRUE
)

#longitudinal
atop_EA_long <- netify::netify(
  dyad_data = atop_EA,
  actor1 = 'country1',
  actor2 = 'country2',
  time = 'year',
  symmetric = TRUE,
  sum_dyads = TRUE,
  diag_to_NA = TRUE,
  missing_to_zero = TRUE
)
  
```


### Netify Parameters

- *dyad_data*: *data object*  
  The data frame containing dyadic data.

- *actor1*: *character*  
  The variable name for the first actor in the dyad.

- *actor2*: *character*  
  The variable name for the second actor in the dyad.

- *time*: *character*  
  The variable name for the time variable if working with longitudinal data.  
  The time variable should be of numeric type. If not provided, the function assumes cross-sectional data.

- *symmetric*: *logical*  
  Whether the network is symmetric (`TRUE`) or directed (`FALSE`).

- *sum_dyads*: *logical*  
  Whether to sum dyadic values (`TRUE`) or keep them separate (`FALSE`).

- *diag_to_NA*: *logical*  
  Whether to set diagonal values to `NA` (`TRUE`) or keep them as zero (`FALSE`).

- *missing_to_zero*: *logical*  
  Whether to replace missing values with zero (`TRUE`) or leave them as `NA` (`FALSE`).


The `netify` function creates a network object that can be used with the `srm` package. The resulting object is a matrix for cross-sectional data and a list of matrices for longitudinal data, where each matrix represents the network at a specific time point. The diagonal values are set to NA, and missing values are replaced with zero.


## SRM Summary Statistics

Now that we have our network objects, we can analyze relationships in the network via the `srm` package. Lets begin by generating summary statistics for the network data. We would look at the rowmeans and actor partner covariance for the cross-sectional data of the network data. 


```{r}

#calculate rowmeans
srm_stats(atop_EA_short, type= "rowmeans")


```


After calculating the row means, we see that countries with higher row means such as China (3.76) and India (2.47) tend to have more intensive security consultation pacts or relationships on average compared to countries with more limited military consultation engagement like Turkmenistan (0.29) or Ukraine (0.35).

```{r}
#calculate actor partner covariance
srm_stats(atop_EA_short, type = "actor_partner_cov")
```


The actor-partner covariance of 0.65 shows a positive relationship that countries with high military consultation obligations are also receiving these consultation agreements in the network.



Lets make use of the `time` argument and look at the colmeans for the longitudinal data of the network. We see we get a list of colmeans for the years 2001, 2013, and 2020 for each actor in the network. 

```{r}

#subset to particular years
yrs= c("2010", "2013", "2015", "2018")

srm_stats(atop_EA_long, type = "colmeans", time = yrs)

```


The column means calculated for particular time periods show a country's average activity with security consultation pacts. The results show China's increasing activity from 0.29 in 2010 to 0.53 in 2018. Similarly, India shows an increase from 0.12 to 0.47. This is in contrast to other countries such as Russia and the USA, which remain stable around 0.24 without much change.


## SRM Effects

We can analyze the actor, partner, and unique effects in the network data using the `srm_effects()` function. The actor effect for observation *i* captures how much that actor deviates from the network average as a sender of ties. It is calculated as:

$$\hat{a}_i = \frac{(n-1)^2}{n(n-2)} X_{i.} + \frac{(n-1)}{n(n-2)} X_{.i} - \frac{n-1}{n-2} \bar{X}$$


where $X_{i.}$ is the row sum for actor *i*, $X_{.i}$ is the column sum, and $\bar{X}$ is the overall network mean.

Similarly, the partner effect shows how much actor *i* tends to receive ties above or below average:

$$\hat{b}_i = \frac{(n-1)^2}{n(n-2)} X_{.i} + \frac{(n-1)}{n(n-2)} X_{i.} - \frac{n-1}{n-2} \bar{X}$$

We would look at the actor effects for the cross-sectional data of the network.


```{r}

EA_actor = srm_effects(atop_EA_short, type = "actor")
print(EA_actor)

```

The actor effects show how much an actor deviates from the network average in security consultation pacts. From this example, we see that China has the highest positive actor effect (+2.92), meaning it engages in security consultation pacts far more than the general network average. Other East Asian countries such as Thailand, Japan, and Korea have negative effects (around -0.52), suggesting they participate in security consultation pacts below the network average.

We can also look at the unique effects as well. The unique dyadic effects is calculated as the following below. Both actor and partner tendencies are removed to isolate relationship-specific components:

$$\hat{g}_{ij} = X_{ij} - \hat{a}_i - \hat{b}_j - \bar{X}$$

We run this from the `srm_effects` function with the `type = "unique"` argument. 


```{r}
EA_unique = srm_effects(atop_EA_short, type = "unique")
```



The dyadic effects show the unique relationship between a pair of actors in military consultations controlling for each actor's general tendencies. The results reveal which bilateral relationships are unusually strong or weak compared to what each country's overall behavior would predict. We will not display the output here because of the size.


Similarly, we can do the same for the longitudinal data. We would look at the partner effects for the years 2001, 2013, and 2020 of the nig_net_long network. The function returns a list of matrices showing partner effects for each year. We will not display the output here beause of the size.  

**NOTE**: For undirected networks, the partner effect is the same as the actor effect. The output is a matrix of the unique effects of each actor on the other actors in the network. 

```{r}
EA_actor_long = srm_effects(atop_EA_long, type = "actor", time = yrs)
```



## Visualization

We might want to visualize the actor, partner, or dyadic effects in the network data. We can do this using the `srm_plot` function. The actor/partner effect plots show actors sorted by the absolute magnitude of their effects, with positive effects in blue and negative effects in red.  Actors with strong effects in either direction (highly positive or highly negative) will appear in the plot. The default number of actors to show is the top 10. Users can change this by setting the `n` argument in the `srm_plot` function.

**Things to note:**

  (1) Users need to put in results from the `srm_effects` function as input to the `srm_plot` function. 
  
  (2) For undirected networks, the partner effect is the same as the actor effect so users are advised to use the `type = actor` argument when plotting if working with symmetric data.


We would take a look at the actor effects for the cross-sectional data of the network. 

```{r}
#include the actor effects from the srm_effects function
srm_plot(EA_actor, type = "actor", n=9)

```

The srm plot for actor effects show positive effects which are in black (right side) and negative effects which are in gray (left side). They show actors that engage in military consultation obligations above or below the network average. China has the highest positive effect, indicating it engages in military consultation obligations significantly more than the network average. Countries such as Turkmenistan have the strongest negative effects, showing they engage in military consultation obligations below the network average



We can also look at the dyadic effects for the cross-sectional data of the network. When users supply dyadic effect, the function returns a heatmap of the dyadic effects in the network data. Similarly, users can set the `n` argument to specify the number of actors to show in the plot and positive effects are in blue and negative effects in red. Default is the top 10 actors. 


```{r, fig.width=8, fig.height=6, dpi=100, dev="png"}

#include the unique effects from the srm_effects function
srm_plot(EA_unique, type = "dyadic", n=12)

```

The srm plot for dyadic effects shows a heatmap of the unique relationship between actor pairs in military consultation obligations after controlling for each actors's general tendencies. Darker cells indicate relationships that are stronger-than-expected, while light gray cells show relationships that are weaker-than-expected, and white cells represent relationships that align with predicted patterns. For example, the USA has strong military consultation relationships with Japan, Korea, and Thailand beyond what their general activity levels would predict. Most cells here are white showing many bilateral relationships on military consultation obligations are similar to what we would expect.



We can also do the same for the longitudinal data. We can take a look at the actor effects for the years 2001, 2013, and 2020 of the nig_net_long network. To do so, we will set `facet=TRUE` for the srm_plot function to compare visualizations across time. Similarly, we can set the `n` argument to specify the number of actors to show in the plot.

```{r}
# include the actor effects from the srm_effects function
srm_plot(EA_actor_long, type = "actor", facet = TRUE, time = yrs, n = 8)

```

The srm plot for longitudinal actor effects shows how actors' military consultation obligation patterns evolved over time. The actor effects are sorted by absolute value, with positive effects in dark bars and negative effects in light gray bars. Across the time periods, China remains the country with the highest positive actor effects (dark bars) showing growth from 2010. This is similar to India. The USA and Russia show reduced positive effects in later years but still maintain relatively stable above-average military consultation obligation patterns. Other countries with gray bars show negative actor effects suggesting they engage in military consultation obligations below what would be expected from the network average.


## Conclusion

The srm package provides a comprehensive toolkit for social relations model analysis, allowing researchers to decompose network effects into actor, partner, and unique components. This vignette demonstrated basic functionality and a real-world application to conflict data.


**References**:

 Dorff, Cassy, and Michael D. Ward. (2013) Networks, Dyads, and the Social Relations Model. Political Science Research Methods 1(2):159–178.
 
Dorff, Cassy, and Minhas Shahryar. (2016). When Do States Say Uncle? Network Dependence and Sanction Compliance. International Interactions 43(4): 563–588.

Leeds, Brett Ashley, Jeffrey M. Ritter, Sara McLaughlin Mitchell, and Andrew G. Long. 2002. Alliance Treaty Obligations and Provisions, 1815-1944. International Interactions 28: 237-260.